Regression Cube Analysis of Cohort Study Data

### 2015-03-25 draft.tex
added Kais revision

\maketitle
\com{Bernhard: Comment on prior occurrences of "Regression Cube" and how we differ}
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{Cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
They can also be used to give recommendations regarding a healthy lifestyle and provide information about wide-spread diseases.
%%
In the standard workflow, physicians translate observations into hypotheses, which are depicted using epidemiological features and then assessed using regression analyses.

An important epidemiological tool for deriving such features are \emph{Cohort studies}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
%%
To reduce any selection bias, subjects are invited at random and without a focus on a specific disease.
%%
The acquired features range from social and lifestyle factors to prior or current diseases and medications as well as medical parameters, such as blood pressure and also comprises non-radiating medical image data. e.g. magnetic resonance imaging (MRI).
%%
Medical image data may be quantified using user-defined landmarks, describing for example shape, volume or diameter of a structure.
%%

To assess the statistical resilience of a hypotheses using regression analyses rarely involves more than three features due to the required subject count.
%%
Due to missing overview techniques, possibly interesting correlations lie within the data, but are not apparent.
%%
Explorative analyses and overview visualizations of the data set as presented by Klemm et al. \cite{Klemm2014VIS} are not tailored to a specific target variable and mostly highlight correlations between variables, which are known to the domain expert (e.g. correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts into overview visualizations, which can either be used for an hypothesis-free analysis or an analysis towards a specific disease or hypotheses.
%%
This is achieved by providing template regression formulas, which are applied to all potential variable combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual variables in the process.

Our contributions are:
\begin{itemize}
	\item An overview visualization technique representing feature interactions using target features.
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
	\item freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	\item analysis of confounding variables by providing comparison views between different formula results
	\item The open and web based approach of the system allows for analysis of any data using the presented method.
\end{itemize}

\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiological research is performed by experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition are extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS} are used to analyze the data using the classical sequential epidemiological workflow, which uses images largely only to communicate results, rather than providing insight.
%%
An alternative data-driven approach is described as follows.

@@

\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
\com{Bernhard: Due to overlaps to VIS'14 paper in this section, this can be shortened. E.g. "More discussion of data types and characteristics can be found in [...]"}
%%
\emph{Cohort studies} are a major epidemiological tool for gathering epidemiological features.
%%
They yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either towards a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric variables, medical parameters, genetic data as well as medical image data.
These features are derived through different modalities, such as questionnaires or medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g. number of born children or period status.

@@

\paragraph{Image Data.}
\com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
%%
Since the Rotterdam study, many modern cohort studies include medical image data.
%%
For ethical reasons, the imaging modalities usually not include ionizing radiation.
%%
The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
%%
These data are hard to analyze as they require segmentation highlighting the structures of interest.
%%
This process is prone to inter and intra-observer variability when carried out manually.
%%
Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
%%
Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).

\paragraph{Confounding Features}
Features influencing the exposure as well as the outcome of a analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounding feature.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and therefore almost all features acquired through cohort studies.
%%
Another important confounder is \emph{Gender}.
%%
Other confounder have to be selected by epidemiologists specific to the investigated condition.

\subsection{Regression Analysis} \label{sec:RegressionAnalysis}
%%
Regression analysis is the most important statistical tool when analyzing epidemiological data.
%%
They are also the foundation of this work.
%%
A regression analysis assesses the influence of one or more (\emph{independent}) features to one target (\emph{dependent}) feature.
%%
The regression model yields a function describing the target feature by weighting the independent features.
%%
Metrics, such as the weightings itself and associated p values and describe the resulting function (the \emph{model}).
%%
R$^2$ values describe the goodness of fit; in other words how well the dependent features describe the target feature.
%%
The value is in the range [0,1], while 1 encodes a perfect fit.

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
\begin{equation}
Dependent \sim Independent_1 + Independent_2 + ... + Independent_n
\label{eq:RegressionNotation}
\end{equation}
%%
The most used regression operators comprise:
%%
\begin{itemize}
	\item $+$ / $-$ inclusion/exclusion of the variable,
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
The type of the target feature restricts the regression type.

\paragraph{Linear Regression for Continuous Target.} The basic type is the linear regression, creating linear weightings for the \emph{independent} features.
%%
The dependent variable has to be of a continuous type.
%%

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models do as opposed to linear models does not allow for extracting a R$^2$ goodness of fit value.
%%
Therefore, pseudo-R$^2$ values are extracted, such as the \emph{Nagelkerke R$^2$}, which mimics the behavior of the R$^2$.
%%
\emph{Nagelkerke R$^2$} cannot be compared to R$^2$ values extracted from linear regression model.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
Tukey already stated in 1977 that data if too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to pairwise assess the influence of independent features towards the target.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

Since we want to use the regression analyses with an overview visualization, we are are interested in all possible combinations of (two or more) independent features describing a target.
%%
We achieve this by introducing dynamic variables $X$, $Y$ and $Z$ into the regression notation.
%%
The method then replaces the dynamic variables with all features in the data set.
%%
In a data set with $n$ (e.g. 100) features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}
would yield $n^2$ (10.000) regression models, describing all possible combinations of two features in the data describing the target $Cancer$.
%%
The major advantage of this notation is that it comes natural to anyone familiar with regression analysis, because it uses the same notation, all operators can be used as before.
%%
This allows for a fast adaptation and in the epidemiological application domain.
%%
With simple adjustments to the formula, different results can be achieved:
%%
\begin{itemize}
	\item $Z \sim X + Y$ calculates all combinations of two features towards all possible target features.
	\item $Cancer \sim X + Y + BodyWeight$ includes the $BodyWeight$ feature into all regression models as feature.
	\item $Cancer \sim X + Y + Z$ calculates all combinations of three features towards the target.
\end{itemize}
%%
The problem with this brute-force approach lies in its complexity.
%%
The number of calculated regression models increases exponentially for each dynamic variable added.
%%
If we assume a data set with 100 features with the formula $Z \sim X + Y$ we calculate 1,000,000 regression models.
%%
When each regression takes about 50~ms calculation, we have to wait roughly 14 hours for the calculation to complete.
%%
Therefore, the computational complexity needs to be reduced.
%%
%\com{Regression formulas are described using standard regression formula notation by introducing variables.}
%%
%\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}

@@

\subsection{Abstracting Regression Results using R$^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/cube_sketch}
 \caption{
 (a) Overview visualization using a heatmap for the formula $Z \sim X + Y$, where $Z$ assumes feature $Age$.
 %%
 The R$^2$ values extracted from the regression formulas depict the goodness of fit and are mapped to color saturation.
 %%
 A saturated color shows a strong correlation.
 %%
 (b) Since $Z$ assumes all features $n$ as given by the formula, it yields $n$ heatmap visualizations.
 %%
 These represent the slices in our cube visualization.
 %%
 The R$^2$ values of each slice voxel is mapped on opacity in the 3D-view, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%% TODO: Cite another paper here!
As described in \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded using color on a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the goodness of fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the R$^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since R$^2$ is scaled between [0,1] it also allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the R$^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g. $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
R$^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of R$^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the R$^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
The goal is to create a overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow}
\com{I'm not satisfied with the structure in this subsection, please proofread it with focus in the train of thought!}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_sketch}
 \caption{
 Workflow of the analysis using \emph{Regression Cubes}.
 %%
 The analysis starts with a set of features given by the data set.
 %%
 Then, the user may specify a formula regarding to specify a hypothesis, or use a predefined formula to start a explorative analysis of the data set.
 %%
 The Regression Cube is then visualized, where the user has either the option to adjust the formula or to derive details on demand on a specific regression.
 %%
 Insights into the data yield either an adjustment of the current formula.
 %%
 By applying a difference view, regression cubes can be compared to each other.
 %%
 The analysis yields insights and hypothesis about feature relations.
 %%
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
The analysis workflow follows the Visual Analytics (VA) Mantra of Keim et al. \cite{Keim}:

@@

\textbf{Show the Important.} The 3D-cube visualization acts as overview over the whole data set.
%%
Regression models with large R$^2$ values can be spotted fast here, steering the attention to the respective slice.

@@

The focus lies on creating a overview visualization and gaining insight into relationships into the data, which triggers further analyses (maybe with other (statistical) tools).

@@

Therefore the system has to be intuitive and comprehensive in order to be adapted by domain experts.

@@

Therefore we have to adapt our design respecting the affordances of web pages.

>>>>

\maketitle
\com{Bernhard: Comment on prior occurrences of "Regression Cube" and how we differ}
%%
Epidemiology aims to characterize health and disease conditions in defined populations (\emph{cohorts}).
%%
Insights about risk factors allow to characterize disease-specific high-risk groups and act as important diagnostic key figures \cite{Fletcher2012}.
%%
They can also be used to give recommendations regarding a healthy lifestyle and provide information about wide-spread diseases.
%%
In the standard workflow, physicians translate observations into hypotheses, which are depicted using epidemiological features and then assessed using regression analyses.

An important epidemiological tool for deriving such features are \emph{cohort studies}, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}.
%%
To reduce any selection bias, subjects are invited at random and without a focus on a specific disease.
%%
The acquired features comprise a wide range.
%
Social and lifestyle factors, prior or current diseases and medications as well as medical parameters, such as blood pressure are gathered.
%%
Medical image data from non-radiating modalities. e.g. from magnetic resonance imaging (MRI) is also acquired in modern studies.
%%
These data may be quantified using user-defined landmarks, describing for example shape, volume or diameter of a structure.
%%

Testing features for association with diseases using regression analysis is one of the most important epidemiological tools.
%%
Using it to assess the statistical resilience of a hypotheses rarely involves \emph{more than three features} due to the required subject count.
%%
Due to missing overview techniques, possibly interesting correlations may be missed.
%%
Explorative analyses and overview visualizations of the data set as presented by Klemm et al. \cite{Klemm2014VIS} are not tailored to a specific target variable and mostly highlight correlations between variables, which are known to the domain expert (e.g. correlation between body size and spine shape).
%%
We incorporate the regression analysis, which is familiar to the domain experts into overview visualizations, which can be used for a hypothesis-free analysis or an analysis towards a specific disease or hypotheses.
%%
This is achieved by providing template regression formulas, which are applied to all potential variable combinations.
%%
Since the notation is familiar to epidemiologists, they can rapidly include their domain knowledge into the analysis process.
%%
Difference views between regression formulas allow to assess the influences of individual variables in the process.

Our contributions are:
\begin{itemize}
	\item An overview visualization technique representing feature interactions using target features.
	\item Visualization techniques, which incorporate overview visualizations of all regression analyses at once as well as details on demand techniques for in-depth investigations of feature relationships.
	\item Freely adjustable regression formulas provide a simple, yet powerful way to adjust the regression analysis to specific hypotheses about the data.
	\item Analysis of confounding variables by providing comparison views between different formula results
	\item The open and web-based approach of the system allows for analysis of any data using the presented method.
\end{itemize}

\section{Epidemiological Background} \label{sec:Background}
%%
This section covers the epidemiological workflow and requirements.
%%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
Epidemiological research is performed by experts from different academic disciplines, such as physicians, statisticians and medical computer scientists focusing on biometrics and image segmentation.
%%
Their goal is to derive disease-specific risk factors by assessing epidemiological features using statistical methods.
%%
As described by Thew et al. \cite{Thew2009} the epidemiological workflow is divided into these different steps:
\begin{itemize}
	\item Clinicians make observations in the daily practice, which are translated into hypotheses.
	\item Epidemiologists compile a list of variables depicting the hypothesis and include confounding variables.
	\item Statisticians assess the association of the derived features towards the investigated disease.
\end{itemize}
%%
Relative risks can be determined if a statistical resilient association of features towards a condition are extracted.
%%
They indicate the per-subject chance of developing the disease.
%%
Reproducibility of results is an epidemiological key requirement and guides all analyses steps.
%%
Statistical programs, such as \texttt{SPSS} are used to analyze the data using the classical sequential epidemiological workflow, which uses images largely only to communicate results, rather than providing insight.
%%
An alternative data-driven approach is described as follows.

@@

\subsection{Epidemiological Data} \label{sec:EpidemiologicalData}
%%
\com{Bernhard: Due to overlaps to VIS'14 paper in this section, this can be shortened. E.g. "More discussion of data types and characteristics can be found in [...]"}
%%
\emph{Cohort studies} are a major epidemiological tool for gathering epidemiological features.
%%
They yield a highly heterogenous and incomplete information space.
%%
Subject data are collected either towards a specific disease or with the widest range possible.
%%
The latter allows the data set to be assessed towards different diseases.
%%
The feature space comprises information about lifestyle, somatometric variables, medical parameters, genetic data as well as medical image data.
These features are derived through different modalities, such as questionnaires or medical examinations or laboratory analyses.
%%
Many features are sparse, such as follow up questions about a medication or treatment of a certain disease.
%%
Other features are exclusive for a sub-group, such as women-specific questions, e.g., number of born children or period status.

@@

\paragraph{Image Data.}
\com{Can probably be cut (reference to Klaus' Paper for more details on this) (see Bernhards comment above)}
%%
Since the Rotterdam study, many modern cohort studies include medical image data.
%%
For ethical reasons, the imaging modalities usually not include ionizing radiation.
%%
The image quality is often inferior to clinical standards, which is a tradeoff between time and cost \cite{Preim2014}.
%%
These data are hard to analyze as they require segmentations to outline structures of interest.
%%
This process is prone to inter- and intra-observer variability when carried out manually.
%%
Automatic or semi-automatic solutions are a remedy for this problem, but are costly and need to be tailored to the structure of interest \cite{Toennies2015}.
%%
Image-derived variables are either dichotomous variables indicating a medical finding by a radiologist or continuous, describing the shape of segmented structures (e.g. volume or diameter).

\paragraph{Confounding Features}
Features influencing the exposure as well as the outcome of an analysis are called confounders.
%%
The analysis model has to be adjusted by normalizing all included features towards the confounding feature.
%%
\emph{Age} is included as confounder in almost any epidemiological analysis, since most diseases (such as different cancer types) are more likely with increasing age.
%%
It also influences the general body condition and thereby almost all features acquired through cohort studies.
%%
Another important confounder is \emph{Gender}.
%%
Other confounders have to be selected by epidemiologists specific to the investigated condition.

\subsection{Regression Analysis} \label{sec:RegressionAnalysis}
%%
Regression analysis is the most important statistical tool when analyzing epidemiological data.
%%
They are also the foundation of this work.
%%
A regression analysis assesses the influence of one or more (\emph{independent}) features to one target (\emph{dependent}) feature.
%%
The regression model yields a function describing the target feature by weighting the independent features.
%%
Different metrics, such as the weightings itself and associated $p$ values describe the resulting function (the \emph{model}).
%%
$R^2$ values describe the goodness of fit; in other words how well the dependent features describe the target feature.
%%
The value is in the range [0,1], where 1 encodes a perfect fit.

\paragraph{Regression Analysis Notation.} Regression formulas are usually denoted as follows:
%%
\com{Kai, + in this case is a regression operator including a feature. The formula should denote any operator (maybe as Var1 omega Var2).}
\begin{equation}
Dependent \sim Independent_1 + ... + Independent_n
%Dependent \sim \sum_{i=1}^n Independent_i
\label{eq:RegressionNotation}
\end{equation}
%%
The most used regression operators comprise:
%%
\begin{itemize}
	\item \fbox{$+,-$} inclusion/exclusion of the variable,
	\item $:$ inclusion of interactions between the variables (e.g. $x:y$),
	\item $*$ inclusion the variables as well as their interactions (e.g. $x*y$)
	\item $|$ (conditioning) inclusion of variable x, given y (e.g. $x|y$)
\end{itemize}
%%
The type of the target feature restricts the regression type.

\paragraph{Linear Regression for Continuous Target.} The basic type is the linear regression, creating a linear map from the space comprising the \emph{independent} features to the \emph{dependent} features.
%%
The \emph{dependent} variable has to be of a continuous type.
%%

\paragraph{Logistic Regression for Dichotomous Target.} Logistic regression implies a dichotomous target variable.
%%
The target is described by fitting a logistic function.
%%
Logistic models as opposed to linear models do not allow for extracting an $R^2$ goodness of fit value.
%%
Therefore, pseudo-$R^2$ values are extracted, such as the \emph{Nagelkerke $R^2$}, which mimics the behavior of the $R^2$.
%%
\emph{Nagelkerke $R^2$} cannot be compared to $R^2$ values extracted from linear regression model.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
The \texttt{SHIP}, located in Northern Germany aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%%
It does not focus on a specific disease, making the data set open for many diseases.
%%
Unique for the \texttt{SHIP} is the acquisition of medical image data per subject.
%%
A second cohort, \texttt{SHIP-TREND} was introduced in 2012.
%%
Data for both cohorts are examined in a 5-year time span.
%%
New parameters are added in each iteration, extending the range of investigated diseases.
%%
For the last acquisition (\texttt{SHIP-2} and \texttt{SHIP-TREND-0}), MRI scans are included into the cohort \cite{Hegenscheid2009, Ivanovska2014}.
%%
However, most examinations occur in all stages and are performed according to the same instructions to enable comparisons over the different stages of the study ("\emph{method loyality}").
%%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
%They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
%Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
%Data are collected without focus on a group of diseases.
%
%This allows to query the data regarding many diseases and conditions.
%
%
\section{Prior and Related Work}
%% https://en.wikipedia.org/wiki/Exploratory_data_analysis#cite_note-Tukey1977-4
Tukey already stated in 1977 that data if too often only analyzed using confirmatory data analysis \cite{Tukey}.
%%
He emphasized the need to use data to derive hypotheses, which can then be tested again.
%%
In this section we present prior and related work trying to achieve this goal.
%%
We put each paper in relation to our work, derive ideas and point out differences, starting with the works closest to ours:
%%
\\\\
The work of Piringer et al. \cite{Piringer}, Chan et al. \cite{Chan}, Angelelli et al. \cite{Angelelli} and Zhang et al. \cite{Zhang2014} is closest to ours regarding different aspects.
%% HyperMoVal: Interactive Visual Validation of Regression Models for Real-Time Simulation
Akin to our approach, Piringer et al. \cite{Piringer} propose methods for visualizing regression analysis results and properties for developing car engines.
%%
Their main goal is to pairwise assess the influence of independent features towards the target.
%%
This is achieved by using a plot matrix displaying models as contours towards the target feature.
%%
They also incorporate 3D visualizations for each pairwise combination, but mainly because it is popular with the engineers in their target domains.
%%
Linked views of model deviations allow to select outlier residuals.
%%
Comparison of models is also included by mapping model values on line-style (e.g. solid for $model_a$ and dashed for $model_b$).
%%
This limits the method to comparing a few models at once.
%%
The plot matrix representation however gets complex and harder to understand with increasing number of independent features.
%%
The main difference is their focus on analyzing one complex model in detail, yielding in extensive plots, while we analyze a large amount of simpler models towards \emph{different} features.
%%
They also focus on metric features, while we process also categorical data.
%%
We share the matrix-arrangement of results depicting relationships between independent features in a model.

@@

Since we want to use the regression analyses with an overview visualization, we are are interested in all possible combinations of (two or more) independent features describing a target.
%%
We achieve this by introducing dynamic variables $X$, $Y$ and $Z$ into the regression notation.
%%
Our method replaces the dynamic variables with all features in the data set.
%%
In a data set with $n$ (e.g. 100) features, the regression formula
\begin{equation}
Cancer \sim X + Y
\end{equation}
would yield $n^2$ (10.000) regression models, describing all possible combinations of two features in the data describing the target $Cancer$.
%%
The major advantage of this notation is that it comes natural to anyone familiar with regression analysis, because it uses the same notation, all operators can be used as before.
%%
This allows for a fast adaptation and in the epidemiological application domain.
%%
With simple adjustments to the formula, different results can be achieved:
%%
\begin{itemize}
	\item $Z \sim X + Y$ calculates all combinations of two features towards all possible target features.
	\item $Cancer \sim X + Y + BodyWeight$ includes the $BodyWeight$ feature into all regression models as feature.
	\item $Cancer \sim X + Y + Z$ calculates all combinations of three features towards the target.
\end{itemize}
%%
The problem with this approach lies in its complexity.
%%
The number of calculated regression models increases exponentially for each dynamic variable added.
%%
If we assume a data set with 100 features with the formula $Z \sim X + Y$ we calculate 1,000,000 regression models.
%%
When each regression takes about 50~ms calculation, we have to wait roughly 14 hours for the calculation to complete.
%%
Therefore, the computational complexity needs to be reduced.
%%
%\com{Regression formulas are described using standard regression formula notation by introducing variables.}
%%
%\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}

@@

\subsection{Abstracting Regression Results using $R^2$}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/cube_sketch}
 \caption{
 (a) Overview visualization using a heatmap for the formula $Z \sim X + Y$, where $Z$ assumes feature $Age$.
 %%
 The $R^2$ values extracted from the regression formulas depict the goodness of fit and are mapped to color saturation.
 %%
 A saturated color shows a strong correlation.
 %%
 (b) Since $Z$ assumes all features $n$ as given by the formula, it yields $n$ heatmap visualizations.
 %%
 These represent the slices in our cube visualization.
 %%
 The $R^2$ values of each slice voxel is mapped on opacity in the 3D-view, reducing the occlusion of other values.
 }
  \label{fig:Cube}
\end{figure}
%%
The goal of an overview visualization is to provide a comprehensive view on the data (raw or using descriptive metrics \cite{Bertini}), which is easy to understand.
%% TODO: Cite another paper here!
As described in our previous work \cite{Klemm2014VIS}, correlation values scaled between 0 (no correlation) and 1 (perfect correlation) can be encoded using color on a mosaic plot.
%%
Regression models are more complex, having many associated describing metrics.
%%
For the \emph{Regression Cube} analysis we are interested in the goodness of fit of the resulting model, which allows to infer about the predictive quality of the independent features included in the model.
%%
As described in Section~\ref{sec:RegressionAnalysis}, the $R^2$ value is the metric allowing for this kind of assessment.
%%
\paragraph{2D (Slice) View.}
%%
Since $R^2$ is scaled between [0,1], it allows for comparison between regression models.
%%
We can apply the same mosaic plot mapping by translating the $R^2$ values to color saturation (Fig.~\ref{fig:Cube} a).
%%
This describes a 2D regression square for dynamic variables $X$ and $Y$ (e.g. $Age \sim X + Y$).
%%

\paragraph{3D (Cube) View.}
%%
Introducing $Z$ creates a 3D \emph{Regression Cube} (Fig.~\ref{fig:Cube} b).
%%
$R^2$ values of each cube entry (\emph{voxel}) is mapped to opacity to reduce the overlap.
%%
The visualization of $R^2$ values derived from different regression cubes (e.g. $Z \sim X + Y$) is misleading, as they can be compared relatively, but not in precise numbers.
%%
Therefore, the $R^2$ results of different regression methods are encoded using different colors (e.g. blue for linear regression and red for logistic regression).
%%
This way, the cube can easily be extended using other regression types.
%%
For cubes having one fixed target feature, such as $Cancer \sim X + Y + Z$ no such encodings are required and the $z$ dimension can be compared directly.
%%
\\\\
The goal is to create an overview visualization for a data set, but on the other hand we also want to incorporate expert knowledge into the visualization by adapting the underlying formulas.
%%
These two approaches do not exclude each other, they rather underline the difference in purpose of the chosen formula.
%%
Different analysis approaches require different starting points using the \emph{Regression Cube}.

\subsection{Analysis Workflow}
%% TODO: Anmerkung Kai einarbeiten Ja, gefällt mir auch nicht so gut. Ich würde zunächst Keims Mantra vorstellen (mit itemize) dann dazu erklären was er damit meint und was es bedeutet. Diese Mantrapunkte nennst du M1-M4 (oder ein anderen Buchstaben). Und dann würde ich bei der Erklärung deines Tools immer wieder darauf verweisen, welcher Punkt nun erfüllt ist (..., which fulfills M2) 
\com{ToDo. Include Kais revision, renaming Matnrapoints and refer to them}
\com{I'm not satisfied with the structure in this subsection, please proofread it with focus in the train of thought!}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_sketch}
 \caption{
 Workflow of the analysis using \emph{Regression Cubes}.
 %%
 The analysis starts with a set of features given by the data set.
 %%
 Then, the user may specify a formula regarding to specify a hypothesis, or use a predefined formula to start a explorative analysis of the data set.
 %%
 The Regression Cube is then visualized, where the user has either the option to adjust the formula or to derive details on demand on a specific regression.
 %%
 Insights into the data yield either an adjustment of the current formula.
 %%
 By applying a difference view, regression cubes can be compared to each other.
 %%
 The analysis yields insights and hypothesis about feature relations.
 %%
 }
  \label{fig:Workflow}
\end{figure}
%%
%\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%\com{Workflow for different analysis approaches.}
%\com{0. Select the Variables.}
%\com{1. Select the Formulas}
%\com{a) Analysis without hypothesis about the data set (z ~ x+y)}
%\com{b) Analysis without hypothesis with confounder (z ~ age)}
%\com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
%\com{d) Use Difference View}
%\com{d) Clarify using confounder (Var ~ x)}
%\com{3. Find interesting pane.}
%\com{4. Get Details on Demand.}
%%
The analysis workflow follows the Visual Analytics (VA) Mantra of Keim et al. \cite{Keim}:

@@

\textbf{Show the Important.} The 3D-cube visualization acts as an overview over the whole data set.
%%
Regression models with large $R^2$ values can be spotted here fast, steering the users attention to the respective slice.

@@

The focus lies on creating an overview visualization and gaining insight into relationships into the data, which triggers further analyses (maybe with other (statistical) tools).

@@

Therefore, the system has to be intuitive and comprehensive in order to be adapted by domain experts.

@@

Therefore, we have to adapt our design respecting the affordances of web pages.
